---
title: "Untitled"
author: "J Lacasa"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries 

```{r message=FALSE, warning=FALSE}
library(tidyverse) # data wrangling, data viz
library(mgcv) # fitting generalized additive models 
library(latex2exp) # math notation in plots
library(ggpubr) # plot aesthetics
```

## Loading the data  

```{r}
# dd_finalproj <- read.csv("../classes/data/dd_finalproj.csv")
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/dd_finalproj.csv"
dd_finalproj <- read.csv(url)
dd_finalproj$doy_f <- factor(dd_finalproj$doy)
```

## Exploratory Data Analysis  

```{r}
dd_finalproj %>% 
  ggplot(aes(doy, fitted))+
  geom_point(aes(y = agb_g, fill = trt), shape = 21)+
  facet_wrap(~species)+
  scale_fill_manual(values = c("#FC4C02", "#008E97"))+
  labs(y = expression(Aboveground~Biomass~(Mg~ha^{-1})), 
       x= "Day of the Year", 
       color = "Treatment", 
       fill = "Treatment")+
  theme_pubclean()+
  theme(aspect.ratio = 1)
```

## Model Fitting  

Firstly, fit a simple model 
$$y_{ijkl} \sim N(\mu_{ijk}, \sigma^2)$$
$$\mu_{ijk}=\beta_0 + \tau_i + \rho_j + \alpha_k + (\tau \rho)_{ij} + (\tau \alpha)_{ik} +  (\rho \alpha)_{ik} +  (\tau \rho \alpha)_{ijk},$$

where $y_{ijkl}$ is the observation of aboveground biomass (in g) for the $i$th treatment, $j$th species, $k$th moment (i.e., time), and $l$th repetition, that arises from a Normal distribution with mean $mu_{ijk}$ and variance $sigma^2$. The $\beta_0$ is the general intercept, $\tau_i$ is the effect of the $i$th treatment, $\rho_j$ is the effect of the $j$th species, and $(\tau \rho)_{ij}$, $(\tau \alpha)_{ik}$, and $(\rho \alpha)_{ik}$ are the two-way interactions between the factors mentioned above, and $(\tau \rho \alpha)_{ijk}$ is the three-way interaction of the factors mentioned above. Note that there is a single variance for all observations.    


```{r message=FALSE, warning=FALSE, include=FALSE}
m1 <- lm(agb_g ~ species*trt*factor(doy), data = dd_finalproj)
# m1 <- lm(agb_g ~ species+trt+doy_f+
#            species:trt+species:doy_f+trt:doy_f+
#            species:trt:doy_f, data = dd_finalproj)
# View(model.matrix(m1))
```

### Model diagnostics  

#### Constant variance  
```{r}
plot(m1$fitted.values, abs(residuals(m1, type = "deviance")), 
     xlab = TeX("$\\hat{y}_i$"), 
     ylab = TeX("|$\\hat{\\epsilon}_i$|"), 
     main = "Residuals versus fitted values", 
     sub = "Note that the errors increase together with the means")
```

```{r}
lm(log(agb_g) ~ species*trt*doy_f, data = dd_finalproj)
```

### Model fitting II  

The first, simpler, model (i.e., `m1`) does not seem to have constant variance. We can follow two routes: (1) transform the response and keep the assumptions, or (2) keep the data as is and change our model assumptions.  

Following the second option, we can model 

$$y_{ijkl} \sim N(\mu_{ijk}, \sigma^2_k)$$
$$\mu_{ijk}=\beta_0 + \tau_i + \rho_j + \alpha_k + (\tau \rho)_{ij} + (\tau \alpha)_{ik} +  (\rho \alpha)_{ik} +  (\tau \rho \alpha)_{ijk},$$  

where $y_{ijkl}$ is the observation of aboveground biomass (in g) for the $i$th treatment, $j$th species, $k$th moment (i.e., time), and $l$th repetition, that arises from a Normal distribution with mean $mu_{ijk}$ and variance at time $k$, $sigma^2_k$. The $\beta_0$ is the general intercept, $\tau_i$ is the effect of the $i$th treatment, $\rho_j$ is the effect of the $j$th species, and $(\tau \rho)_{ij}$, $(\tau \alpha)_{ik}$, and $(\rho \alpha)_{ik}$ are the two-way interactions between the factors mentioned above, and $(\tau \rho \alpha)_{ijk}$ is the three-way interaction of the factors mentioned above. Note that the variance is a function of time.  


```{r}
m2 <- gam(list(agb_g ~ species*trt*factor(doy),
               ~ doy),
          family = gaulss(),
          data = dd_finalproj)
m2_fitted <- predict(m2)[,1]
m2_residuals <- residuals(m2, type = "deviance")
```

### Model diagnostics II  

#### Constant variance  

```{r}
plot(m2_fitted, abs(m2_residuals), 
     xlab = TeX("$\\hat{y}_i$"), 
     ylab = TeX("$\\hat{\\epsilon}_i$"), 
     main = "Residuals versus fitted values", 
     sub = "The errors no longer increase together with the means, 
     but there seems to be different dispersions for different species")
```


### Model fitting III    

The second, model (i.e., `m2`) fixes the issue of constant variance. However, the variance seems to be different depending on the groups.  

We can thus further model $\sigma^2$:  

$$y_{ijkl} \sim N(\mu_{ijk}, \sigma^2_{jk})$$
$$\mu_{ijk}=\beta_0 + \tau_i + \rho_j + \alpha_k + (\tau \rho)_{ij} + (\tau \alpha)_{ik} +  (\rho \alpha)_{ik} +  (\tau \rho \alpha)_{ijk},$$  

where $y_{ijkl}$ is the observation of aboveground biomass (in g) for the $i$th treatment, $j$th species, $k$th moment (i.e., time), and $l$th repetition, that arises from a Normal distribution with mean $mu_{ijk}$ and variance at time $k$ for species $j$, $sigma^2_{jk}$. The $\beta_0$ is the general intercept, $\tau_i$ is the effect of the $i$th treatment, $\rho_j$ is the effect of the $j$th species, and $(\tau \rho)_{ij}$, $(\tau \alpha)_{ik}$, and $(\rho \alpha)_{ik}$ are the two-way interactions between the factors mentioned above, and $(\tau \rho \alpha)_{ijk}$ is the three-way interaction of the factors mentioned above. Note that the variance is a function of time.  

```{r}
m3 <- gam(list(agb_g ~ species*trt*factor(doy),
               ~ doy + species),
          family = gaulss(),
          data = dd_finalproj)
m3_fitted <- predict(m3)[,1]
m3_residuals <- residuals(m3, type = "deviance")
```

### Model diagnostics III  

#### Constant variance  

```{r}
plot(m3_fitted, abs(m3_residuals), 
     xlab = TeX("$\\hat{y}_i$"), 
     ylab = TeX("$\\hat{\\epsilon}_i$"), 
     main = "Residuals versus fitted values", 
     sub = "These look OK")
```

#### Normal distribution    

```{r}
car::qqPlot(m2_residuals)
car::qqPlot(m3_residuals)
```

```{r}
hist(m2_residuals)
hist(m3_residuals)
```

### AIC, BIC 


```{r}
left_join(AIC(m1,m2,m3) %>% rownames_to_column("model"), 
          BIC(m1,m2,m3) %>% rownames_to_column("model"))
```


### Other checks  

```{r}
gam.check(m3)
```

### Inference  

#### Summary  
```{r}
summary(m3)
```

#### ANOVA  
```{r}
anova(m3)
```

#### Means, Mean comparisons  
```{r message=FALSE, warning=FALSE}
means <- emmeans::emmeans(m3,
                          ~trt:species, at = list("doy" = c(237)))

mean_comparisons <- multcomp::cld(means,
                                  level = 0.05,
                                  adjust = "none",
                                  decreasing = TRUE,
                                  Letters = letters) %>%
  mutate(.group = trimws(.group))

mean_comparisons %>% 
  transmute(species, trt,SE=round(SE,2),
            emmean = round(emmean, 2), group = .group,
            lower.CL = round(lower.CL,2), upper.CL = round(upper.CL,2))
```


```{r message=FALSE, warning=FALSE}
ggplot(mean_comparisons, aes(emmean, species))+
  geom_errorbarh(aes(xmin = emmean-SE, xmax = emmean + SE, color = trt), height = 0.1)+
  geom_point(aes(color = trt) , position = "dodge")+
  scale_fill_manual(values = c("#FC4C02", "#008E97"))+
  scale_color_manual(values = c("#FC4C02", "#008E97"))+
  theme_pubclean()+
  labs(x = expression(Aboveground~Biomass~(Mg~ha^{-1})), 
       y = "Species",
       color = "Treatment", 
       fill = "Treatment")+
  theme(aspect.ratio = 1, 
        legend.position = "bottom")

# ggsave("finalproj_figure1.png", width = 5, height = 5)
```

```{r}
means2 <- emmeans::emmeans(m3,
                          ~trt:species, at = list("doy" = c(155)))

mean_comparisons2 <- multcomp::cld(means2,
                                  level = 0.05,
                                  adjust = "none",
                                  decreasing = TRUE,
                                  Letters = letters) %>%
  mutate(.group = trimws(.group))

ggplot(mean_comparisons2, aes(emmean, species))+
  geom_errorbarh(aes(xmin = emmean-SE, xmax = emmean + SE, color = trt), height = 0.1)+
  geom_point(aes(color = trt) , position = "dodge")+
  scale_fill_manual(values = c("#FC4C02", "#008E97"))+
  scale_color_manual(values = c("#FC4C02", "#008E97"))+
  theme_pubclean()+
  labs(x = expression(Aboveground~Biomass~(Mg~ha^{-1})), 
       y = "Species",
       color = "Treatment", 
       fill = "Treatment")+
  theme(aspect.ratio = 1, 
        legend.position = "bottom")

means3 <- emmeans::emmeans(m3,
                          ~trt:species, at = list("doy" = c(194)))

mean_comparisons3 <- multcomp::cld(means3,
                                  level = 0.05,
                                  adjust = "none",
                                  decreasing = TRUE,
                                  Letters = letters) %>%
  mutate(.group = trimws(.group))
ggplot(mean_comparisons3, aes(emmean, species))+
  geom_errorbarh(aes(xmin = emmean-SE, xmax = emmean + SE, color = trt), height = 0.1)+
  geom_point(aes(color = trt) , position = "dodge")+
  scale_fill_manual(values = c("#FC4C02", "#008E97"))+
  scale_color_manual(values = c("#FC4C02", "#008E97"))+
  theme_pubclean()+
  labs(x = expression(Aboveground~Biomass~(Mg~ha^{-1})), 
       y = "Species",
       color = "Treatment", 
       fill = "Treatment")+
  theme(aspect.ratio = 1, 
        legend.position = "bottom")

```

```{r}
plot_data <- expand.grid(doy = unique(dd_finalproj$doy), 
                         species = c("A", "B", "C", "D", "E"), 
                         trt = c("ctrl", "flood"))

plot_data <- bind_cols(plot_data, 
                       predict(m3, newdata = plot_data, type = "link"))

plot_data <- plot_data %>% 
  rename(fitted = `...4`, sd = `...5`) %>% 
  mutate(sd = .01+exp(sd))

plot_data %>% 
  ggplot(aes(doy, fitted))+
  geom_line(aes(color = trt))+
  geom_ribbon(aes(ymin = fitted-sd*1.96, ymax = fitted+sd*1.96, fill = trt), alpha = .4)+
  geom_point(aes(y = agb_g, color = trt), data = dd_finalproj, shape = 21, alpha = .3)+
  facet_wrap(~species)+
  scale_fill_manual(values = c("#FC4C02", "#008E97"))+
  scale_color_manual(values = c("#FC4C02", "#008E97"))+
  labs(y = expression(Aboveground~Biomass~(Mg~ha^{-1})), 
       x= "Day of the Year", 
       color = "Treatment", 
       fill = "Treatment")+
  theme_pubclean()+
  theme(aspect.ratio = 1, 
        strip.background = element_rect(fill = NA, color = "black"))
```


### Bonus: What if we hadn't changed our model?  

This is just a demonstration of what happens when we design a model that does not describe our data generating process well. This does not necessarily have to go in your project. 

```{r}
plot_data <- expand.grid(doy = unique(dd_finalproj$doy), 
                         species = c("A", "B", "C", "D", "E"), 
                         trt = c("ctrl", "flood"))

plot_data <- bind_cols(plot_data, 
                       predict(m1, newdata = plot_data, interval = "prediction"))

plot_data %>% 
  ggplot(aes(doy, fit))+
  geom_line(aes(color = trt))+
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = trt), alpha = .4)+
  geom_point(aes(y = agb_g, color = trt), data = dd_finalproj, shape = 21, alpha = .3)+
  facet_wrap(~species)+
  scale_fill_manual(values = c("#FC4C02", "#008E97"))+
  scale_color_manual(values = c("#FC4C02", "#008E97"))+
  labs(y = expression(Aboveground~Biomass~(Mg~ha^{-1})), 
       x= "Day of the Year", 
       color = "Treatment", 
       fill = "Treatment")+
  theme_pubclean()+
  theme(aspect.ratio = 1, 
        strip.background = element_rect(fill = NA, color = "black"))
```

