# Modeling designed experiments in the presence of spatial variability  
October 8th 

## Announcements 

- Zoom classes on 10/13 and 10/15. 
- Assignment 4 due today midnight. 

## Spatial variability  

- Most commonly found in field experiments. 
- Whiteboard: going from the normal iid model to a spatial dependence model. 

### Spatial patterns 

Consider the model
$$\mathbf{y} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma}),$$
where $\boldsymbol{\Sigma}$ is the variance-covariance model. 
In the presence of spatial patterns, there are parametric and non-parametric tools to account for spatial dependence. 

Sometimes, we can clearly define the shape of the spatial dependence pattern. For example:

**Matérn covariance function**

$$C(d; \sigma^2, \phi, \kappa) = \sigma^2 \frac{(h/\phi)^\kappa K_\kappa (h/\phi)}{2^{\kappa-1} \Gamma(\kappa)}$$

**Gaussian covariance function**

$$C(d; \sigma^2, \phi) = \sigma^2 \exp \left\{ - \frac{d^2}{2\phi^2} \right\} $$

**Exponential covariance function**

$$C(d; \sigma^2, \phi, \kappa) = \sigma^2 \exp\{-(d/\phi)^\kappa\}$$

### Applied example  

Yield data from an advanced Nebraska Intrastate Nursery (NIN) breeding trial conducted at
Alliance, Nebraska, in 1988/89.

```{r}
library(tidyverse)
library(agridat)
library(ggpubr)
data <- drop_na(stroup.nin)

data %>% 
  ggplot(aes(col, row))+
  geom_tile(aes(fill = rep), color = "black")+
  theme_pubr()+
  labs(fill = "Block")+
  theme(axis.line = element_blank())
```

The statistical model:

$$y_{ij}\sim N(\mu_{ij}, \sigma^2), \\
\mu_{ij} = \eta_{ij} = \eta_0 + G_i + b_j, \\
b_j \sim N(0, \sigma^2_b).$$

Assumptions: 

- iid, normal residuals 

```{r}
library(lme4)
m <- lmer(yield ~ gen + (1|rep), data = stroup.nin)
data$res <- resid(m)
```

```{r}
data %>% 
  ggplot(aes(col, row))+
  geom_tile(aes(fill = res), color = "black")+
  theme_pubr()+
  scico::scale_fill_scico()+
  theme(axis.line = element_blank())+
  labs(fill = expression(Residual~(bu~ac^{-1})))
```

```{r}
as.data.frame(emmeans(m, ~ gen)) %>% 
  arrange(-emmean) %>% 
  head()
```


```{r}
library(glmmTMB)

data$coordinate <- numFactor(data$col, data$row)
data$group <- factor(1) 

m_spatial <- glmmTMB(yield ~ gen + (1|rep) + mat(coordinate + 0 | group), data = data)
```

```{r}
data$resid_spatial <- resid(m_spatial)

data %>% 
  ggplot(aes(col, row))+
  geom_tile(aes(fill = resid_spatial), color = "black")+
  theme_pubr()+
  scico::scale_fill_scico()+
  theme(axis.line = element_blank())+
  labs(fill = expression(Residual~(bu~ac^{-1})))
```

```{r}
as.data.frame(emmeans(m_spatial, ~ gen)) %>% 
  arrange(-emmean) %>% 
  head()
```

